---
title: 'Exercise 5: WAIC'
author: "Jason Doll"
date: "June 8, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Objective
This exercise will demonstrate how to do Bayesian model comparison using Watanabe's Information Criterion (WAIC)

## Background
In many applications, it is neccessary to compare two or more models to determine which model "best" describes the data. Although there are many methods, this exercise will demonstrate how to modify a Stan model to calculate WAIC. You are given the R code to generate simulated data and two basic Stan programs. The first file, "EX5_WAIC_LR.stan", estimates paramters of a linear regression model and the second file, "Ex5_WAIC_ME.stan", esimtates parameters of a mixed effect model. The "EX5_WAIC_LR.stan" model you will fit to observations i is:

\[
\begin{aligned}
\text{Model:} \hspace{1cm} \text{y}_i &= \alpha + \beta \text{X}_i + \epsilon_i\\
\hspace{1cm} \epsilon_i &\sim \text{Normal}(0, \sigma)\\
\text{Priors:} \hspace{1cm} \alpha &\sim \text{Normal}(0, 100)\\
\hspace{1cm} \beta &\sim \text{Normal}(0, 100)\\
\sigma &\sim \text{half-cauchy}(0, 5)
\end{aligned}
\]

The "EX5_WAIC_ME.stan" model you will fit to observations i and group j is:

\[
\begin{aligned}
\text{Model:} \hspace{1cm} \text{y}_i &= \alpha_{j[i]} + \beta \text{X}_i + \epsilon_i\\
\hspace{1cm} \epsilon_i &\sim \text{Normal}(0, \sigma)\\
\text{Priors:}\hspace{1cm} \alpha_j &\sim \text{Normal}(\mu, \tau)\\
\hspace{1cm} \mu &\sim \text{Normal}(0, 100)\\
\hspace{1cm} \tau &\sim \text{half-cauchy}(0, 5)\\
\hspace{1cm} \beta &\sim \text{Normal}(0, 100)\\
\sigma &\sim \text{half-cauchy}(0, 5)
\end{aligned}
\]

You will modify the two Stan programs to generate a new quantity to hold the log-likelihood. This parameter will be extracted and WAIC calculate using the loo package. 

### R packages required for this exercise
1. rstan
2. shinystan
3. loo

## Directions
Add a "generated quantities block" with the following code to the two existing files "Ex5_WAIC_LR.stan" and "Ex6_WAIC_ME.stan". Note, you MUST use the quantity "log_lik".

## Stan code for centered parameterization
```{r, engine="stan",output.var="Ex1", eval=F, echo=T}
generated quantities{
    vector[n] log_lik;
    
    for ( i in 1:n ) {
        log_lik[i] = normal_lpdf( y[i] | alpha + beta * x[i], sigma_y);
    }
}

```

## Stan code for non-centered parameterization
```{r, engine="stan",output.var="Ex1", eval=F, echo=T}
generated quantities{
    vector[n] log_lik;
    
    for ( i in 1:n ) {
        log_lik[i] = normal_lpdf( y[i] | gpmu[gp[i]] + beta * x[i], sigma_y);
    }
}

```

### R Code
After modifying both .stan files to include the "generated quantities" block above, open "Ex5_WAIC.R" and run lines 1 through the lines below. This should be around line 100-107.
```{r, eval=F, echo=T}
#Noncentered Example without Divergent transitions
non_Center_WAIC_ME = stan(file = 'Ex5_WAIC_ME.stan',
                  data = dataList , 
                  init = initslst,
                  chains = nchains,
                  iter = 1000 ,
                  warmup = 500 , 
                  thin = 2,
                  control = list(adapt_delta=0.99, max_treedepth=15))
```

```{r , echo=F, include=F, eval=T}

#load libraries
library(rstan)
library(shinystan)
library(loo)

#Set working directory to source file locations
#In Rstudio 
#This directory must have all data files and Stan model code needed.
#If you receive an error, manually set the working directory
#setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

#clear workspace
rm(list=ls())

#set random number seed for consistent data
set.seed(14898) 
#Generate data
gp=6 #Number of groups/individuals
n=100 #Number of measurements per group/individual

subject = rep (1:gp, each = n)  #vector of code for group/individual
mu_true = 50 # True mean
sigma_y = 30 # Population SD
sigma_mu = 40 # Group SD
beta = 10     #slope
x = runif(n,0,100)     #values of covariate, year

mu_ind = rnorm(gp, mu_true, sigma_mu)
y = rnorm (gp * n, mean = mu_ind[subject] + beta *x , sd = sigma_y)

data1 = data.frame(y, subject, x = x)

#Package data, initialization values, and send to Stan
#Data in dataList(can use any name you want) must match values in Stan data block

#specify number of chains, used to initialize values and specify chains
nchains = 3

dataList = list(
  n= (gp*n),
  y = data1$y,
  x=data1$x,
  gp = data1$subject,
  ngps = gp
)
#Initialize values
#convergence can be improved by setting reasonable starting values
#i.e, range of observations from 1-20, don't intialize mean at 100000
#Use different starting values for each chain
initslst = lapply(1:nchains,function(i) {
  list(
    sigma=runif(1,1,10),
    alpha=runif(1,min(y),max(y)),
    beta=runif(1,-10,10)
  )
})

#Centered Example with Divergent Transitions
WAIC_FE = stan(file = 'Ex5_WAIC_LR_final.stan',
             data = dataList , 
             init = initslst,
             chains = nchains,
             iter = 1000 ,
             warmup = 500 , 
             thin = 2,
             control = list(adapt_delta=0.99))



#specify number of chains, used to initialize values and specify chains
nchains = 3
dataList = list(
  n= (gp*n),
  y = data1$y,
  x=data1$x,
  gp = data1$subject,
  ngps = gp
)

#Initialize values
#convergence can be improved by setting reasonable starting values
#i.e, range of observations from 1-20, don't intialize mean at 100000
#Use different starting values for each chain
initslst = lapply(1:nchains,function(i) {
  list(
    sigma_y=runif(1,1,10),
    sigma_mu=runif(1,1,10),
    mu=runif(1,min(y),max(y)),
    beta=runif(1,-10,10),
    gpmu_raw = rnorm(gp,0,1)
  )
})

#Noncentered Example without Divergent transitions
non_Center_WAIC_ME = stan(file = 'Ex5_WAIC_ME_final.stan',
                  data = dataList , 
                  init = initslst,
                  chains = nchains,
                  iter = 1000 ,
                  warmup = 500 , 
                  thin = 2,
                  control = list(adapt_delta=0.99, max_treedepth=15))


```
After executing both models the first step to calculate WAIC is to extract the log_lik parameter from both stan objects.

```{r, eval=T, echo=T}
log_likFE = extract_log_lik(WAIC_FE)
log_likME = extract_log_lik(non_Center_WAIC_ME)
```

Finally, calculate the WAIC value
```{r}
waic(log_likFE)
waic(log_likME)
```

Your results should show that the mixed effects model (one that produced the log_likME object above) will have the lowest WAIC value.
