---
title: 'Exercise 6: Iris data, Linear Regression'
author: "Jason Doll"
date: "June 8, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Objective
This exercise will use skills obtained in previous exercises to create the linear regression model.

## Background
In this example, we will use the Iris data set that is included in base R. You will write a stan program to estimate parameters of a linear regression model to predict sepal width from sepal length. The R code to initialize workspace, read data, and package and send the data with initial values to Stan are provided in the "Ex6_ext_lr.R" file located in the "Ex6_Iris" folder. This is a self driven exercise, review the R code then create the .stan file. The model you will fit to individual observations i is:

\[
\begin{aligned}
\text{Model:} \hspace{1cm} \text{y}_i &= \alpha + \beta \text{X}_i + \epsilon_i\\
\hspace{1cm} \epsilon_i &\sim \text{Normal}(0, \sigma)\\
\text{Priors:} \hspace{1cm} \alpha &\sim \text{Normal}(0, 100)\\
\hspace{1cm} \beta &\sim \text{Normal}(0, 100)\\
\sigma &\sim \text{half-cauchy}(0, 5)
\end{aligned}
\]

Where Y = sepal width and X = sepal length

### R packages required for this exercise
1. rstan

## Directions
Create a new text file and save it as "Ex6_est_lr.stan" in the "Ex6_Iris" folder. Using the code from previous exercises, create the neccessary stan model. The R code and final figures are below.

## R Code

Load library, clear workspace, and load data
```{r, results=FALSE}
#####CSTAT Workshop###############
#####Exercise 5###################
#####Analysis of Iris data########
#####File provided by instructor##
##################################

#install/load the rstan package
require(rstan)

#Set working directory to source file locations
#In Rstudio 
#This directory must have all data files and Stan model code needed.
#If you receive an error, manually set the working directory
#setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

#clear workspace
rm(list=ls())

#Generate simulated data
#Load Iris data

iris = iris

#specify number of chains, used to initialize values and specify chains
nchains = 3

# Specify data:
dataList = list(
  'n'=nrow(iris),
  'x'=iris$Sepal.Length,
  'y'=iris$Sepal.Width
)
```

Initialize paramter values and send everything to stan.

```{r, results=FALSE}
#Initialize values
#convergence can be improved by setting reasonable starting values
#i.e, range of observations from 1-20, don't intialize mean at 100000
#Use different starting values for each chain
initslst <- lapply(1:nchains,function(i) {
  list(
    alpha = rnorm(1,0,1),
    beta = rnorm(1,0,1),
    sigma=runif(1,1,10)
  )
})

#send everything to Stan
fit2 <- stan(file = 'Ex6_est_lr.stan',
             data = dataList , 
             init = initslst,
             chains = nchains,
             iter = 1000 ,
             warmup = 500 , 
             thin = 1 )
```

View traceplots and pararmeter estimates
```{r}
#View traceplots
traceplot(fit2)
#view results
fit2
```

Extract paramters and plot histograms
```{r}
#extract results
est_alpha=rstan::extract(fit2,"alpha")$alpha
est_beta=rstan::extract(fit2,"beta")$beta
est_sd=rstan::extract(fit2,"sigma")$sigma

#plot results
par(mfrow=c(1,3))
hist(est_alpha,breaks=20);
hist(est_beta,breaks=20);
hist(est_sd,breaks=20);
par(mfrow=c(1,1))
```

