---
title: "Exercise 1: Estimating Mean and SD"
author: "Jason Doll"
date: "June 8, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Objective

In this execise, you will be introduce the Stan programing language by estimating the mean and standard deviation (SD) of a simulated dataset. You will use concepts describe during the workshop to create your first Stan program

## Background

Data used in this exercise will be randomly generate from a normal distribution with fixed mean and SD. The R code to initialize workspace, generate data, and package and send the data with initial values to Stan are provided in the "Ex1_mean.R" file located in the "Ex1_mean" folder. The instructor will review this R file with you. Your task will be to create the Stan file using the Stan code provided below. You are encouraged to type rather than copy/paste the code into a new .stan document. The model you will fit to individual observations i is:

\[
\begin{aligned}
\text{Model:} \hspace{1cm} \text{y}_i &= \mu + \epsilon_i\\
\hspace{1cm} \epsilon_i &\sim \text{Normal}(0, \sigma)\\
\text{Priors:} \hspace{1cm} \mu &\sim \text{Normal}(0, 100)\\
\sigma &\sim \text{half-cauchy}(0, 5)
\end{aligned}
\]

### R packages required for this exercise
1. rstan

## Directions

Create a new text file and save it as "Ex1_est_mean_sd.stan" in the "Ex1_mean" folder. Enter the Stan code below in the new .stan file. 

## Stan code

```{r, engine="stan",output.var="Ex1", eval=F, echo=T}
data{
  int<lower=0> n; //number of obserations
  vector[n] y;    //observations as a vector
}

parameters {
  real<lower=0> sigma;  //standard deviation to be estimated
  real mu;              //Mean to be estimated
}

model {
  
  //reference priors
  sigma ~ cauchy(0,5);
  mu ~ normal(0,100);
  
  //likelihood, loop through number of observations
  for(i in 1:n){
    y[i] ~ normal(mu, sigma);
  }
  
}

```

### R Code

Open "Ex1_mean.R" and run lines 1 through the lines:

```{r, eval=F, echo=T}
#send everything to Stan
fit1 <- stan(file = 'Ex1_est_mean_sd.stan',
             data = dataList , 
             init = initslst,
             chains = nchains,
             iter = 1000 ,
             warmup = 500 , 
             thin = 1 )
```

Traceplots, parameter estimates and plots of results can be viewed using the code below (note this code is provided):

```{r, eval=F, echo=T}
#View traceplots
traceplot(fit1)
#view results
fit1

#extract results
est_mean=rstan::extract(fit1,"mu")$mu
est_sd=rstan::extract(fit1,"sigma")$sigma

#plot results
par(mfrow=c(1,2))
hist(est_mean,breaks=50);abline(v=mu,lwd=5);
hist(est_sd,breaks=50);abline(v=sd,lwd=5);
par(mfrow=c(1,1))
```
