---
title: 'Exercise 2: Linear Regression'
author: "Jason Doll"
date: "June 8, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Objective
This execise will expand on exercise 1 by introducing linear regression as an extension of the simple model of the mean using the Stan programing language. This exercise will provide practice at coding regression models in Stan.

## Background
In this example, we will randomly generate data for a response variable, y, with known parameters and predictors, x. The R code to initialize workspace, generate data, and package and send the data with initial values to Stan are provided in the "Ex2_LReg.R" file located in the "Ex2_LReg" folder. The instructor will review this R file with you. Your task will be to create the Stan file using the code provided below. You are encouraged to type rather than copy/paste the code into a new .stan document. The model you will fit to individual observations i is:

\[
\begin{aligned}
\text{Model:} \hspace{1cm} \text{y}_i &= \alpha + \beta \text{X}_i + \epsilon_i\\
\hspace{1cm} \epsilon_i &\sim \text{Normal}(0, \sigma)\\
\text{Priors:} \hspace{1cm} \alpha &\sim \text{Normal}(0, 100)\\
\hspace{1cm} \beta &\sim \text{Normal}(0, 100)\\
\sigma &\sim \text{half-cauchy}(0, 5)
\end{aligned}
\]

### R packages required for this exercise
1. rstan

## Directions
Create a new text file and save it as "Ex2_est_lr.stan" in the "Ex1_mean" folder. Enter the Stan code below in the new .stan file. 

## Stan code
```{r, engine="stan",output.var="Ex1", eval=F, echo=T}
data { 
  int<lower=0> n;  //number of observations
  vector[n] x;     //observed x values, predictors
  vector[n] y;     //observed y values, response
}

parameters { 
  real<lower=0> sigma; //standard deviation
  real alpha;          //y-intercept
  real beta;           //slope
} 

model { 
  //reference priors
  alpha ~ normal(0,100);
  beta ~ normal(0,100);
  sigma ~ cauchy(0,5);

  //likelihood, loop through number of observations
  for(i in 1:n){
    y[i] ~ normal(alpha + beta * x[i], sigma); 
  }
}

```

### R Code
Open "Ex1_mean.R" and run lines 1 through the lines:

```{r, eval=F, echo=T}
#send everything to Stan
fit2 <- stan(file = 'Ex2_est_lr.stan',
             data = dataList , 
             init = initslst,
             chains = nchains,
             iter = 1000 ,
             warmup = 500 , 
             thin = 1 )

```

Traceplots, parameter estimates and plots of results can be viewed using the code below (note this code is provided):

```{r, eval=F, echo=T}
#View traceplots
traceplot(fit2)
#view results
fit2

#extract results
est_alpha=rstan::extract(fit2,"alpha")$alpha
est_beta=rstan::extract(fit2,"beta")$beta
est_sd=rstan::extract(fit2,"sigma")$sigma

#plot results
par(mfrow=c(1,3))
hist(est_alpha,breaks=50);abline(v=a,lwd=5);
hist(est_beta,breaks=50);abline(v=b,lwd=5);
hist(est_sd,breaks=50);abline(v=sd,lwd=5);
par(mfrow=c(1,1))

```

